---
title: "Getting started with the efts R package"
author: "Jean-Michel Perraud"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with the efts R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE
)
library(efts)
```

## Overview

**efts** is an R package to access ensemble forecast time series stored (EFTS) in netCDF format. It offers 
convenient functions to access time series, hiding the bug-prone details of netCDF array manipulations. 

EFTS netCDF data sets follow the schema described at [this location](https://github.com/jmp75/efts/docs/netcdf_for_water_forecasting.md) at the time of writing.

The package comes with API documentation, as well as the present vignette. You can access both navigating via

```
?efts
```

## Creating EFTS netCDF files

Perhaps as an unusual introduction, we will learn to write a EFTS netCDF data set prior to reading it.

Typical EFTS data is multidimensional, and the netCDF schema reflects this. To create a new data set, 
you need to define these dimensions. To create the information about the main time axis, you can use the 
`create_time_info` helper function in the package. The following command creates an hourly data set over two days

```{r}
timeAxisStart <- ISOdate(year=2010, month=08, day=01, hour = 12, min = 0, sec = 0, tz = "UTC")
timeDimInfo <- create_time_info(from=timeAxisStart, n=48, tStep='hours since', tStepDelta=1L)
str(timeDimInfo)
```

Other dimensions are easier to create. Variables are more involved; one way to define several variables is 
via a data frame. For instance, you can start from a stub such as created by the following:

```{r}
variable_names <- c('rain_fcast_ens','pet_fcast_ens')
varDef <- create_variable_definition_dataframe(variable_names=variable_names, longNames = rep('synthetic data', 2))
str(varDef)
```

Do note the column names of the data frame `varDef`; the package is picky about these, to comply with the EFTS netCDF data schema.

Let's create a data set with 2 point stations, 3 ensembles and four time step forecast lead.

```{r}
stationIds <- c(123,456)
nEns <- 3
nLead <- 4
fname <- tempfile() # or something you prefer.
default_var_attributes <- create_var_attribute_definition()
snc <- create_efts(fname, timeDimInfo, varDef, stationIds, nc_attributes=default_var_attributes, lead_length=nLead, ensemble_length=nEns)
```

The default `print` method for this object snc is the same output as objects of class `ncdf4`

```{r}
snc
```

`snc` is a type of object that few R aficionados are aware of; this is a reference class. Without entering 
into unnecessary technical details, this is mostly a design choice done to achieve better memory usage and performance in some contexts. 

The following command displays the main characteristics of this reference object, of class `EftsDataSet`
Note that you should really use the methods, and not access directly the fields.

```{r}
str(snc, max.level=2)
```

You can get to the documentation page for the methods in this class with:

```
?efts::EftsDataSet
```

Our EFTS data set object is ready to be populated with data. Let's create synthetic data. 
The basic idea of the object's methods is to offer intuitive and concise means to get/set ensemble 
of forecast time series. A method can be called in a syntax that may be unfamiliar to most R users, 
but similar to most object oriented languages: `theObject$theMethodName(someArgumentsIfAny)`, 
for instance `snc$get_time_dim()` in our EFTS data set, to retrieve its time axis.

TODO: test, document and demonstrate missing value handling.

```{r}
set.seed(42)
td <- snc$get_time_dim()
for (i in 1:length(td)) {
	for (station in stationIds) {
		rain <- 6 * rnorm(nEns*nLead)
		rain <- matrix(pmax(as.numeric(rain), 0), nrow=nLead) # nEns replicates of a forecast of length nLead
		pet <- 6.0 + rnorm(nEns*nLead)
		pet <- matrix(pmax(as.numeric(pet), 0), nrow=nLead)
		dtime = td[i]
		snc$put_ensemble_forecasts(rain, variable_name = variable_names[1], identifier = station, start_time = dtime) 
		snc$put_ensemble_forecasts(pet,  variable_name = variable_names[2], identifier = station, start_time = dtime) 
	}
}
```

## Reading data

Now we can demonstrate how to retrieve data. If you had closed the previous EFTS data set object `snc` and deleted the variable, you'd reopen it for reading with the following command:

```{r, eval=FALSE}
if (!exists('snc')) snc <- open_efts(fname)
```

You get the ensemble forecast for a variable (pet in this case) for a point in time with the following command

```{r}
td <- snc$get_time_dim()
timeStamp <- td[5] # for instance
d <- snc$get_ensemble_forecasts(variable_names[2], stationIds[1], start_time=timeStamp)
str(d)
```

The object returned is of class `xts`. The package **xts** has been chosen as the default time series structure for **efts**. The rationale is empirical: previous experience with the **xts** in conjunction with **plyr** (note to self: possibly **dplyr** in the future) showed good performance to calculate hydrologic statistics on ensemble of time series.

To quickly visualize it (if the data is not too big), using the function `plot.zoo` has the advantage of stacking the series:

```{r}
zoo::plot.zoo(d)
```

As of version 0.5-1, there is a convenience function to retrieve the UTC offset of the units of time dimension. Here there is none, hence a duration of zero returned:

```{r}
paste0( "UTC offset as a string: ", snc$get_utc_offset())
paste0( "UTC offset as a lubridate duration: ", snc$get_utc_offset(as_string=FALSE))
```

Close the data set with the following commands:

```{r}
snc$close()
rm(snc)
```

This vignette creates by default a temporary file. This shall be cleaned up by default with:

```{r}
if(file.exists(fname)) file.remove(fname)
```
